A Reproducible Workflow with RMarkdown, Git, Make & Docker

### 2019-08-20 rr-flow.Rmd
correct grammary and spelling

Reproducible Research has become the minimal standard for scientific work [@hardwicketome.DataAvailabilityReusability] in recent years. The following tutorial aims to outline a workflow that guarantees not only a reasonable standard of reproducibility but also decreases the burden of maintaining software and code for the researcher and her colleagues. The authors' believe that integrating tools from software engineering in the process of scientific reporting to enhance reproducibility is crucial and give guidelines for the practitioner to use them. These tools include software for literate programming, version control, dependency tracking, and containerization. If implemented, such workflow enables the researcher to transparently create, dynamicly render and puplicly share reports, which she and her colleages can trust to reproduce later, regardless of operating system or software version.

@@

While it is widely recognized that these tools enhance reproducibility and there are several guidelines for using them (Literate Programming: @ruleTenSimpleRules2019; Version Control: @barbaHardRoadReproducibility2016; Dependency Management: @askrenUsingMakeReproducible2016; Containerization: @clyburne-sherinComputationalReproducibilityContainers2018) Within the community of users of the R Programming Language for each purpose a specific software solutions is clearly the most popular. For literate programming, `rmarkdown`, for version control `git`, for dependency management `make` and for container `docker`. Each of them serves a meta-scientific valuable goal (reproducibility) and increases the producitivity of the researcher. Each of these software solutions are extremly mighty, so in order to master them fully often years of practice are required. Luckily the minimal valuable subset needed to ensure reproducibility can be learned with little practice.

The authors believe that for sufficient chances of reproducibility, good documentation is crucial, since it anwers the question "What steps have to be taken to reproduce the results?". The most precise documentation is computer code and the most productive computer code is code that runs automaticly without human interaction. These principles are what guides the use of `rmarkdown`, `git`, `make`, `docker` and there interaction.

The first step towards reproducibility is to have a R script or R markdown, that can be executed on the own computer without error. We assume that the reader is familar with R and has the ability to archive that for a simple analysis. A next step is to make sure that all files relevant, can be moved to another computer. To archive that it is important that all files are within one folder (and subfolders within it) and all paths are relative to that folder. A nice solution to that problem are [Rstudio Projects](https://r4ds.had.co.nz/workflow-projects.html) and/or the `here` package.

@@

Reproducibility at a basic level assumes that results are the same if neither script nor data have changed. It is often not trivial to find out whether anything has changed and if so to "go back in time." `git` enables you to do both. A good mental model for `git` is that it takes a sequence of snapshots of all files it is supposed to track. In the language of git, these snapshots are "commits".  A commit represents a complete copy of the state of the files when added to a commit. Each commit has a unique identifier (a hash). Going back to one state is as easy as finding the hash of the commit. This can be made even easier by tagging specific commits e.g., as "submission," "preprint," "publication." This collection of commits is called "repository," which is almost the same as your Project.

@@

So to keep track of all changes, you only need to use `git add` & `git commit` or `git commit -a` to do both. These commands need to be executed in the `terminal`, which you can access from within Rstudio (`Shift + Alt + R`), but if you use Git with Rstudio, you also get a graphical user interface and quite likely you won't need the terminal more then once per project.

@@

Even when you have obtained a version of the project, and you can ensure that this version is unchanged, you may not know how to reproduce the results, because it is unclear what has to be run in which order. This job gets comfortable with `make`, since it allows you to create (computational) recipes too (re-)create files. A downside of `make` is, that the installation on Windows is not straigtfowrward. However if `docker` (see later sections) is availible this is not an issue, since `make` ships with (almost) every `docker` image.

@@

If you have followed the above workflow a fellow researcher is three commands away from fully reproducing your analysis:

@@

However, this relies on the crucial assumption that your computational environments are sufficiently compatible e.g., everything needed is installed (R + Packages) at a compatible version of the software.

`docker` is a tool that allows to precisely state what computational environment an analysis relies on, in a way that can automatically be recreated on most operating systems (Windows, OS X & Linux). It does that, by creating a virtual computer on which a series of commands (e.g., installing software) is executed. The resulting state of the virtual computer is saved in what is called an "image." This image can be started and execute commands, e.g., running an R-script or `make`. A running instance of an image is called a container. An image can be transferred and executed on any machine that has docker installed. Regardless of the machine the container is executed on, from the perspective of a programm running inside a container, it always looks identical. There are essential differences to a traditional virtual machine, but thinking about a docker container as a computer running inside your computer is still a plausible mental model. The most important advantage over traditional virtual machines is that container are lightweight, meaning they start fast and don't need much storage. Docker achieves that by reusing large parts of the hosts operating system (how much is varying widely between Linux, OS X & Windows).

The following example show why researcher might wan't to invest time into docker. While the R programming language is considered largely stable and much effort is put into backwarts comptibility, sometimes even basic functions like `sample()` change their behavior from one version to another.

@@

The usual expectation is that this code will deliver the same pseudo random five numbers regardless of operating system or R-Version (due to `set.seed()`). Using docker we can start an image which contains the R-Version 3.5.0 and execute the code there.

@@

However in an image with a more recent version of R (3.6.0) this results in another sample.

@@

Note that this is intended behavior und the result of a [bugfix](https://bugs.r-project.org/bugzilla/show_bug.cgi?id=17494) implemented beginning of R 3.6.0.

Such change will prevent many analysis from beeing replicated, if they contain e.g., multiple imputation, bootstrapping, simulations studies, graphics with random jitter, bayesian estimations, or similar technics which involve random sampling. However seldom does a analysis depend only on R, often a considerable number of packages is required, and they will most likly more often introduce breaking changes with updates. The whole endouver of reproducibility is therefore at stake, everytime a update is rolled out. In order to prevent that `docker` replicates the  computational environment of an analysis exactly. However we do not intend to advocate that software should not be updated, it should just be documented and easily replicated which version was used. Quite to the contrary with solutions like docker it gets easier then ever to savely update to new versions, by just bumping the R Version number of the docker image the analysis is relying on.

This convinience is possible to the efforts of the [rocker project](https://github.com/rocker-org/rocker) [@boettigerIntroductionRockerDocker2017a] which provides docker images preconfigured with an R-installation of a fixed R-version. These rely on MRAN [@revolutionanalyticsReproducibilityUsingFixed2019], a mirror of CRAN, the repository from which R-packages are installed from, which is fixed to the last date the R-version of the image was the most recent R-version. Building upon these rocker-images researcher can build there own docker-images with the required R-packages. The rocker project provides also images which do include RStudio (`rocker/rstudio`), the `tidyverse` package (`rocker/tidyverse`) and the `rmarkdown` package with Latex (`rocker/verse`). Since the describes workflow relies on `rmarkdown` we suggest using the `rocker/verse` image (which also contains `rstudio` & `tidyverse`). These images are stored on [dockerhub](https://hub.docker.com) [@dockerinc.DockerHub2019].

To extent the rocker images e.g. with other packages a Dockerfile (a File actually named `Dockerfile`) is needed, the docker image for this article is based on the following Dockerfile:

@@

The `FROM` statement specifies which docker image to use, in this case, the `rocker/verse` image which is tagged with 3.6.1 (referring to the R version 3.6.1). The `RUN` statement describes a command to execute, in this case, to run an R-script `install2.r` which is available on all rocker images, to install the packages `pacman`,  `here` & `pander`. A Dockerfile allows more than one `RUN` statement, executing any available command. Those `RUN` statements can install dependencies that are not an R package e.g., other programming languages like python or Matlab. The `WORKDIR` statement is not strictly necessary, but saves writing the working directory out every time a command is issued to docker. The command `docker build -t image-name` creates an image named `image-name` from the Dockerfile in the project. Following there are two principle ways to share a docker image, either by sharing the Dockerfile which created the image or by sharing the image itself e.g. through [dockerhub](https://hub.docker.com). While both ways garuantee a replicable computational environment, sharing the Dockerfile is more transparent and changes to it are tracked by git.

@@

The `--rm` flag means to not save the state of the docker after the command finishes in an image. The `-it` flag means to accept inputs and return outputs to the terminal. This is especially usefull when using `R` as a command, because then you have acces to the standard R-terminal.

>>>>

Reproducible Research has become the minimal standard for scientific work [@hardwicketome.DataAvailabilityReusability] in recent years. The following tutorial aims to outline a workflow that guarantees not only a reasonable standard of reproducibility but also decreases the burden of maintaining software and code for the researcher and her colleagues. The authors' believe that integrating tools from software engineering in the process of scientific reporting to enhance reproducibility is crucial and give guidelines for the practitioner to use them. These tools include software for literate programming, version control, dependency tracking, and containerization. If implemented, such workflow enables the researcher to transparently create, dynamically render, and publicly share reports, which she and her colleagues can trust to reproduce later, regardless of operating system or software version.

@@

The scientific community widely recognizes that these tools enhance reproducibility, and provides guidelines for their use (Literate Programming: @ruleTenSimpleRules2019; Version Control: @barbaHardRoadReproducibility2016; Dependency Management: @askrenUsingMakeReproducible2016; Containerization: @clyburne-sherinComputationalReproducibilityContainers2018). However, only a workflow using them in unison allows confidence in reproducing a scientific report. Within the community of users of the R Programming Language for each purpose, a specific software solution is the most popular. For literate programming, `rmarkdown`, for version control `git`, for dependency management `make` and for container `docker`. Each of them serves a meta-scientific valuable goal (reproducibility) and increases the productivity of the researcher. Each of these software solutions is exceptionally mighty, so to master them thoroughly often years of practice are required. Luckily the researcher can learn the minimal valuable subset needed to ensure reproducibility with little practice.

The authors believe that for sufficient chances of reproducibility, proper documentation is crucial since it answers the question "What steps have to be taken to reproduce the results?". The most precise documentation is computer code, and the most productive computer code is code that runs automatically without human interaction. These principles are what guides the use of `rmarkdown`, `git`, `make`, `docker` and their interaction.

The first step towards reproducibility is to have an R-script or R-markdown, that runs on the own computer without error. A process, we assume, that is familiar to the reader. A next step is to make sure that all files relevant for the analysis can be moved to another computer. To archive that it is essential that all files are within one folder (and subfolders within it) and all paths are relative to that folder. A friendly solution to that problem are [Rstudio Projects](https://r4ds.had.co.nz/workflow-projects.html) and the `here` package.

@@

Reproducibility at a basic level assumes that results are the same if neither script nor data have changed. It is often not trivial to find out whether anything has changed and if so to "go back in time." `git` enables you to do both. A good mental model for `git` is that it takes a sequence of snapshots of all files it is supposed to track. In the language of git, these snapshots are "commits".  A commit represents a complete copy of the state of the files when added to a commit. Each commit has a unique identifier (a hash). Going back to one state is as easy as finding the hash of the commit. Finding a specific hash can be made even easier by tagging specific commits, e.g., as "submission," "preprint," "publication." This collection of commits is called "repository," which is almost the same as your Project.

@@

So to keep track of all changes, you only need to use `git add` & `git commit` or `git commit -a` to do both. These commands need to be executed in the `terminal`, which you can access from within Rstudio (`Shift + Alt + R`).  Rstudio also offers a graphical user interface for Git and quite likely this suffice for almost every interaction with Git.

@@

Even when you have obtained a version of the Project, and you can ensure that this version is unchanged, you may not know how to reproduce the results, because it is unclear what has to be run in which order. This job gets comfortable with `make`, since it allows you to create (computational) recipes too (re-)create files. A downside of `make` is, that the installation on Windows is not straightforward. However, if `docker` (see later sections) is available this is not an issue, since `make` ships with (almost) every `docker` image.

@@

If you have followed the above workflow, a fellow researcher is three commands away from fully reproducing your analysis:

@@

However, this relies on the crucial assumption that your computational environments are sufficiently compatible, e.g., everything needed is installed (R + Packages) at a compatible version of the software.

`docker` is a tool that allows to precisely state what computational environment an analysis relies on and recreates this environment automatically on most operating systems (Windows, OS X & Linux). `docker` does that, by setting up a virtual computer on which it executes a series of commands (e.g., installing software). It then saves the resulting state of the virtual computer in what is called an "image." This image can be started and execute commands, e.g., running an R-script or `make`. A running instance of an image is called a container. An image can be transferred and executed on any machine that has docker installed. Unregarding the machine executing the container, from the perspective of a program running inside the container, it always looks identical. There are essential differences to a traditional virtual machine, but thinking about a docker container as a computer running inside your computer is still a plausible mental model. The most important advantage over traditional virtual machines is that container are lightweight, meaning they start fast and don't need much storage. Docker achieves that by reusing large parts of the hosts operating system (how much is varying widely between Linux, OS X & Windows).

While the R programming language is considered stable and much effort is put into backward compatibility, sometimes even basic functions like `sample()` change their behavior from one version to another.

@@

The usual expectation is that this code delivers the same pseudo-random five numbers regardless of the operating system or R-Version (due to `set.seed()`). Using docker, we can start an image which contains the R-Version 3.5.0, and execute the code there.

@@

Surprisingly in an image with a more recent version of R (3.6.0), this results in another sample:

@@

Note that this is the intended behavior and the result of a [bugfix](https://bugs.r-project.org/bugzilla/show_bug.cgi?id=17494) implemented from R 3.6.0 upwards.

Such change prevents many reports from beeing replicated, if they contain, e.g., multiple imputations, bootstrapping, simulations studies, graphics with random jitter, bayesian estimations, or similar technics which involve random sampling. However seldom does an analysis depend only on R, often a considerable number of packages is required, and they will most likely more often introduce breaking changes with updates. The whole endeavor of reproducibility is therefore at stake, every time an update is rolled out. To prevent that `docker` replicates the computational environment of analysis exactly. We do not intend to advocate that software should not be updated; the used software version should just be documented and easily replicated. Quite to the contrary with solutions like docker, it gets more comfortable than ever to safely update to new versions, by just bumping the R Version number of the docker image.

This convenience is possible to the efforts of the [rocker project](https://github.com/rocker-org/rocker) [@boettigerIntroductionRockerDocker2017a] which provides docker images preconfigured with an R-installation of a fixed R-version. These rely on MRAN [@revolutionanalyticsReproducibilityUsingFixed2019], a repository for R-Packages, which is fixed to the last date the R-version of the image was the most recent. Building upon these rocker-images researchers can build there own docker-images with the required R-packages. The rocker project also provides images which do include RStudio (`rocker/rstudio`), the `tidyverse` package (`rocker/tidyverse`) and the `rmarkdown` package with Latex (`rocker/verse`). Since the describes workflow relies on `rmarkdown` we suggest using the `rocker/verse` image (which also contains `rstudio` & `tidyverse`). These images are stored on [dockerhub](https://hub.docker.com) [@dockerinc.DockerHub2019].

The basic rocker images are extendable by a Dockerfile, e.g., the basis for docker image for this article is the following Dockerfile:

@@

The `FROM` statement specifies which docker image to use, in this case, the `rocker/verse` image with the tag 3.6.1 (referring to the R version 3.6.1). The `RUN` statement describes a command to execute, in this case, to run an R-script `install2.r` which is available on all rocker images, to install the packages `pacman`,  `here` & `pander`. A Dockerfile allows more than one `RUN` statement, executing any available command. Those `RUN` statements can install dependencies that are not an R package, e.g., other programming languages like python or Matlab. The `WORKDIR` statement is not strictly necessary but saves time spent writing the working directory. The command `docker build -t image-name` creates an image named `image-name` from the Dockerfile in the project. Following there are two principal ways to share a docker image, either by sharing the Dockerfile which created the image or by sharing the image itself e.g., through [dockerhub](https://hub.docker.com). While both ways guarantee a replicable computational environment, sharing the Dockerfile is more transparent, and git tracks changes to it.

@@

The `--rm` flag means to not save the state of the docker after the command finishes in an image. The `-it` flag means to accept inputs and return outputs to the terminal. Interacting with docker this way is especially useful when using `R` as a command because then you have access to the standard R-terminal.
